---
title: "Multilevel Modeling of Arbellay et al 2017"
author: "Song S. Qian"
date: "October 14, 2017"
output: pdf_document
---

# Front Matters

Function for loading/installing packages:
```{r, prompt=TRUE}
## load/install packages
packages<-function(x, repos="http://cran.r-project.org", ...){
  x<-as.character(match.call()[[2]])
  if (!require(x,character.only=TRUE)){
    install.packages(pkgs=x, repos=repos, ...)
    require(x,character.only=TRUE)
  }
}
```

Set up working directory:

```{r}
base <- getwd()
dataDIR <- paste(base, "data", sep="/")
## put your data set in the Data subdirectory
plotDIR <- paste(base, "Figs", sep="/")
## put created figures in the Figures subdirectory
setwd(base)
```

Load needed packages:
```{r, include=TRUE, echo=TRUE, message=FALSE, results='hide'}
packages(lattice)
packages(arm)
packages(reshape2)
packages(tikzDevice)
```

# Reading and Processing Data

```{r, tidy=TRUE}
tp282 <- read.csv(paste(dataDIR, "ArbellayAll.csv", sep="/"))
## removing log-values
tp282<-tp282[,-grep("(LOG)$", names(tp282))]
## creating events
tp282$Events <- cut(tp282$YEAR, breaks=c(1885, 1890, 1909, 1939, 1946, 1955, 1967), label=1:6)
tp282$trt <- 2
tp282$trt[is.element(tp282$YEAR, c(1886,1887,1906,1907,1935,1936,1943,1944,1952,1953,1961,1962))] <- 1
tp_melt<-melt(tp282, id.vars=c("SAMPLE","YEAR","Events","trt"))
```

# Multilevel Model

Details of the model are in Qian (2016) [Environmental and Ecological Statistics with R (2nd Ed.), Chapman and Hall/CRC Press], Specifically, Section 10.3 (Multilevel ANOVA) and graphics code used here are similar to the code in Section 10.6 (Multilevel GLM).

```{r, tidy=TRUE}
tp_lmer1 <- lmer(log(value) ~ 1+(1|Events)+(1|trt)+(1|variable), data=tp_melt)
summary(tp_lmer1) ## This model assumes additive effects for all three factors

tp_melt$EvTp <- paste(tp_melt$Events, tp_melt$trt)
tp_lmer2 <- lmer(log(value) ~ 1+(1|EvTp)+(1|variable), data=tp_melt)
summary(tp_lmer2)

tp_melt$MeEvTp <- paste(tp_melt$variable, tp_melt$Events, tp_melt$trt)
tp_lmer3 <- lmer(log(value) ~ 1+(1|MeEvTp), data=tp_melt)
summary(tp_lmer3) ## interactions among all three factors (implicitly)
```

# Processing output

The three models all perform well.  The estimated residual variances are all close to 0.36.  Using the full model, we summarize the result graphically.

```{r, tidy=TRUE}
est <- as.data.frame(matrix(unlist(strsplit(row.names(coef(tp_lmer3)[[1]]), split=" ")), ncol=3, byrow=T))
names(est) <- c("method", "events", "treatment")
est$treatment<-factor(est$treatment, labels = c("after", "before"))
est$estimate <- coef(tp_lmer3)[[1]][,1]
est$se <- sqrt(se.fixef(tp_lmer3)^2 + se.ranef(tp_lmer3)[[1]][,1]^2)
est$se2 <- se.ranef(tp_lmer3)[[1]][,1]
```

## Plotting results 
The function `my.panel` is written to add CIs to the estimated effects.
```{r, tidy=TRUE, fig.align='center', fig.width=7.5, fig.height=3.5}
my.panel <- function(x, y, subscripts, group.number, col, pch, se, ...){
    myjitter <- c(-0.1,0.1)
    panel.dotplot(x,as.numeric(y)+myjitter[group.number],
                  cex=0.5, pch=pch, col=col)
##    panel.grid()
    panel.segments(est$estimate[subscripts]-est$se2[subscripts]*2,
                   as.numeric(y)+myjitter[group.number],
                   est$estimate[subscripts]+est$se2[subscripts]*2,
                   as.numeric(y)+myjitter[group.number],
                   col=col, 
                   pch=pch)
}
est_low<-tapply(est$estimate-2*est$se2, est$method, min)
est_up <- tapply(est$estimate+2*est$se2, est$method, max)
est_range <- rbind(est_low, est_up)

##trellis.device(pdf, file=paste(plotDIR, "compareCL.pdf", sep="/"), height=3.5, width=7.5, color = T)
key <- simpleKey(levels(est$treatment), space="right")
dotplot(events~estimate|method, data=est,
        key=key, 
        groups=treatment,
        col=key$points$col,
        pch=key$points$pch,
        layout=c(4,1),
        scales=list(x=list(relation="free")),
        panel=my.panel, xlab="log effects", ylab="events",
##        scales=list(x=list(alternating=T)),
        xlim=list(est_range[,1], est_range[,2], est_range[,3], est_range[,4]))
##dev.off()
##trellis.device(pdf, file=paste(plotDIR, "compareBW.pdf", sep="/"), height=3.5, width=7.5, color=F)

key <- simpleKey(levels(est$treatment), space="right")
key$points$pch=c(1,4)
dotplot(events~estimate|method, data=est,
        key=key,
        groups=treatment,
        col=key$points$col,
        pch=key$points$pch,
        layout=c(4,1),
        scales=list(x=list(relation="free")),
        panel=my.panel, xlab="log effects", ylab="events",
##        scales=list(x=list(alternating=T)),
        xlim=list(est_range[,1], est_range[,2], est_range[,3], est_range[,4]))
##dev.off()
```
